Warming Degrees and Flowering Trees

### 2025-04-03 paper/paper.qmd
fixed diagrams in eda

Before the advent of modern technology, people understood themselves, their environments, and their cultures through the cycle of life and death. The Japanese tradition of *hanami* (loosely translated as 'flower-viewing') is an important cultural tradition that honors this perspective, being associated with the flowering birth of cherry blossom trees in the late spring. In the city of Kyoto in particular, local newspapers have recorded the dates of tree flowering since the late 19th century, but researchers like @Aono2010 have been able to estimate flowering dates as far back as the year 812 CE, by combing through historical records such as diaries, poetry and other written documents describing *hanami* parties throughout the ages. However, rising global temperatures create an anomaly, highlighting the need to understand the historical and modern links between temperature and cherry blossom bloom timing. Without it, cherished traditions like *hanami* may struggle to adapt to climate shifts, risking their disappearance as a unique Japanese custom. To this end, work like @Allen2014's analyse and relatively modern blooming data over a 29-year period, but the key weakness of this is a lack of perspective. Given that their data only covers a small part of the entire recorded history of cherry blossom blooming, their analysis and subsequent model have a key weakness - they do not take into account large scale trends of blooming and their relationship to the change in climate that has happened in the last 150 years, which @Aono2010's data can illustrate.

We seek to answer the question of whether warmer temperatures lead to earlier blooming times. To do this, we analyse and model the nearly 1200-year record compiled by @Aono2010 in tandem with modern data compiled by the Japanese Meteorological Agency (JMA), collected and compiled together by @Cookson. @sec-lit-review does a broad review of existing work surrounding cherry blossoms, their significance, and efforts to understand their behaviour related to climate. @sec-data covers the data sources, methodology, and conducts an exploratory analysis of the dataset. @sec-model proposes a model of cherry blossom flowering day as a function of mean March temperature in Kyoto and lists its results. @sec-discussion critically examines the data and results, comes to several conclusions about the dataset, and proposes new areas of exploration. This analysis highlights several conclusions. First, temperature changes will have consequential effects for wildlife that function around the timing of blooming *sakura*. Next, social celebrations such as *hanami* will be impacted, changing the timing of a once-constant annual event. Finally, there is a need for more a more comprehensive inquiry into the determinants of cherry blossom blooming in Japan more generally, involving more covariates, so that more robust predictions of *sakura* blooming in Kyoto can be made. By gathering more data to make better predictions, the city of Kyoto will better be able to weather climate change and bridge their cherished traditions into the future.

@@

```{python}
#| label: tbl-historical-data
#| tbl-cap: "Historical Cherry Blossom Blooming"

# Rename columns for better readability
historic_bloom_renamed = historic_bloom.rename(columns={
    'year': 'Year',
    'flower_date': 'Bloom Date',
    'flower_doy': 'Day of Year',
    'avg_temp_march_c': 'March Temp (°C)'
})
# Display the table without .style.set_caption()
display(historic_bloom_renamed.head())
```

@@

```{python}
#| label: tbl-modern-data
#| tbl-cap: "Modern Cherry Blossom Blooming"

@@

In our data exploration, the `matplotlib` [@matplotlib] and `seaborn` packages were used to visualize and explore the data. The datasets were first explored chronologically, beginning with the historical data and ending with the modern data.

HISTORICAL DATA

```{python}
#| label: historic-temp-yr-scatter
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

# Scatterplot of historic temp vs years 
sns.set(style="whitegrid")
plt.figure(figsize=(10, 6))

# Scatter plot with pink color
plt.scatter(historic['year'], historic['avg_temp_march_c'], alpha=0.7, color='#FF9ECD', edgecolor='#E75480', s=50)

# Add moving average line in light blue
historic['moving_avg'] = historic['avg_temp_march_c'].rolling(window=10).mean()
plt.plot(historic['year'], historic['moving_avg'], color='#4BA3C3', linewidth=2.5, label='10-Year Moving Avg')

# Add reference line for mean temperature
mean_temp = historic['avg_temp_march_c'].mean()
plt.axhline(y=mean_temp, color='#87CEFA', linestyle='--', alpha=0.8, 
            label=f'Historical Mean ({mean_temp:.1f}°C)')

# Customize the plot with pink/blue theme
plt.title('Historic Temperature vs Year', fontsize=16, color='#E75480')
plt.xlabel('Year', fontsize=12, color='#4BA3C3')
plt.ylabel('Temperature (°C)', fontsize=12, color='#4BA3C3')
plt.xticks(rotation=45, color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.grid(True, alpha=0.3, color='#E0E0E0')

# Add pink/blue background touches
plt.gca().set_facecolor('#F8F8FF')  # Light background
plt.gcf().set_facecolor('#F8F8FF')

plt.tight_layout()
plt.show()
```
This scatterplot displays historical annual temperature records (in °C) from the year 800 CE to 2000 CE. Each pink dot represents the recorded temperature for a given year, while the solid teal line represents a 10-year moving average, smoothing out year-to-year fluctuations to highlight longer-term trends. The dashed horizontal line represents the historical mean temperature of 6.2°C. Notably, the data show considerable fluctuations over the centuries, with a relatively stable pattern through much of the pre-industrial era. However, a sharp and continuous rise in average temperature is observed beginning in the late 19th to early 20th century, suggesting significant warming recently. This trend aligns with broader patterns of climate change and may have implications for phenological events such as the timing of sakura (cherry blossom) flowering.

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

# Scatterplot of historic year vs flowering day
plt.figure(figsize=(12, 7))

@@

# Create scatter plot with pink color
plt.scatter(filtered_historic['year'], filtered_historic['flower_doy'], alpha=0.7, color='#FF9ECD', 
            edgecolor='#E75480', s=50, label='Flowering Day')

# Add moving average line with light blue color
window_size = 10  # 10-year moving average
filtered_historic['moving_avg_doy'] = filtered_historic['flower_doy'].rolling(window=window_size).mean()
plt.plot(filtered_historic['year'], filtered_historic['moving_avg_doy'], color='#4BA3C3', linewidth=2.5, 
         label=f'{window_size}-Year Moving Avg')

# Add reference line for mean with light blue
plt.axhline(y=mean_doy, color='#87CEFA', linestyle='--', alpha=0.8, 
            label=f'Historical Mean ({mean_doy:.1f}, April 13-14)')

# Customize the plot with pink/blue theme
plt.title('Cherry Blossom Flowering Day Throughout History', fontsize=16, color='#E75480')
plt.xlabel('Year', fontsize=12, color='#4BA3C3')
plt.ylabel('Day of Year (DOY)', fontsize=12, color='#4BA3C3')
plt.grid(True, alpha=0.3, color='#E0E0E0')
plt.xticks(color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.legend(framealpha=0.7)

# Create a secondary y-axis with date labels
ax = plt.gca()
ax2 = ax.twinx()

# Get the current y-limits from the primary axis
y_min, y_max = ax.get_ylim()

# Create properly aligned DOY-to-date mapping
# In a non-leap year:
# DOY 90 = March 31
# DOY 91 = April 1
# DOY 121 = May 1

doy_values = np.arange(int(y_min), int(y_max) + 1, 5)  # Create ticks every 5 days
date_labels = []

for doy in doy_values:
    if doy < 90:
        # DOYs below 90 (March 31)
        day_in_mar = doy - 59  # 59 = DOYs in Jan (31) + Feb (28)
        date_labels.append(f'Mar {day_in_mar}')
    elif doy < 120:
        # DOYs for April (90-119)
        day_in_apr = doy - 90 + 1
        date_labels.append(f'Apr {day_in_apr}')
    elif doy < 151:
        # DOYs for May (120-150)
        day_in_may = doy - 120 + 1
        date_labels.append(f'May {day_in_may}')
    else:
        # DOYs for June and beyond
        day_in_jun = doy - 151 + 1
        date_labels.append(f'Jun {day_in_jun}')

# Set the properly aligned ticks and labels
ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3', fontsize=14)
ax2.tick_params(axis='y', colors='#4F86C6')

plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, axis='y')

# Add pink/blue background touches
plt.gca().set_facecolor('#F8F8FF')  # Light background
plt.gcf().set_facecolor('#F8F8FF')

plt.tight_layout()
plt.show()
```
This graph showcases the annual full flowering dates of sakura from 800 to 2000 CE, plotted as day-of-year (DOY) values, where higher values indicate later blooming in the calendar year. Each pink dot corresponds to an individual year's observed full bloom date, and the solid teal line shows a 10-year moving average; the dashed horizontal line indicates the historical mean, which is 104.6. The data suggest that, while flowering dates have fluctuated across centuries, a notable trend toward earlier blooming began in the late 19th and 20th centuries. This shift coincides with the period of modern climate warming seen in the corresponding temperature graph, supporting the hypothesis that rising spring temperatures are advancing sakura bloom time.

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

# Violin Plot

# Create a figure
plt.figure(figsize=(14, 8))

# Define correct century order with 9th century first
century_order = ['9th', '10th', '11th', '12th', '13th', '14th', '15th', 
                '16th', '17th', '18th', '19th', '20th', '21st']

# Create a custom color palette that transitions from light blue to pink
start_color = '#ADD8E6'  # Light blue
end_color = '#FFC0CB'    # Light pink

# Generate a custom colormap
n_colors = len(century_order)
colors = [start_color, end_color]
cmap_name = 'blue_to_pink'
cm = LinearSegmentedColormap.from_list(cmap_name, colors, N=n_colors)

# Get the colors for each century
custom_palette = {}
for i, century in enumerate(century_order):
    custom_palette[century] = cm(i/float(n_colors-1))

# Plot the violin plot with the custom order and palette
sns.violinplot(x='century_label', y='flower_doy', data=historic, 
               palette=custom_palette, inner='quartile', order=century_order)

# Add individual data points with small jitter
sns.stripplot(x='century_label', y='flower_doy', data=historic, 
              color='#FF9ECD', alpha=0.3, jitter=True, size=4, order=century_order)

# Add reference line for mean flowering day (assuming mean_doy is defined elsewhere)
mean_doy = historic['flower_doy'].mean()  # Calculate if not already defined
plt.axhline(y=mean_doy, color='#E75480', linestyle='--', alpha=0.8, 
            label=f'Historical Mean ({mean_doy:.1f}, April 14)')

# Enhance visuals
plt.title('Cherry Blossom Flowering Day Distribution by Century', fontsize=18, color='#E75480')
plt.xlabel('Century', fontsize=14, color='#4BA3C3')
plt.ylabel('Day of Year (DOY)', fontsize=14, color='#4BA3C3')
plt.xticks(rotation=45, color='#4F86C6')
plt.yticks(color='#4F86C6')

# Create a secondary y-axis with date labels
ax = plt.gca()
ax2 = ax.twinx()

# Get the current y-limits from the primary axis
y_min, y_max = ax.get_ylim()

# Create properly aligned DOY-to-date mapping
# In a non-leap year:
# DOY 90 = March 31
# DOY 91 = April 1
# DOY 121 = May 1

doy_values = np.arange(int(y_min), int(y_max) + 1, 5)  # Create ticks every 5 days
date_labels = []

for doy in doy_values:
    if doy < 90:
        # DOYs below 90 (March 31)
        day_in_mar = doy - 59  # 59 = DOYs in Jan (31) + Feb (28)
        date_labels.append(f'Mar {day_in_mar}')
    elif doy < 120:
        # DOYs for April (90-119)
        day_in_apr = doy - 90 + 1
        date_labels.append(f'Apr {day_in_apr}')
    elif doy < 151:
        # DOYs for May (120-150)
        day_in_may = doy - 120 + 1
        date_labels.append(f'May {day_in_may}')
    else:
        # DOYs for June and beyond
        day_in_jun = doy - 151 + 1
        date_labels.append(f'Jun {day_in_jun}')

# Set the properly aligned ticks and labels
ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3', fontsize=14)
ax2.tick_params(axis='y', colors='#4F86C6')

plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, axis='y')

# Set background color
plt.gca().set_facecolor('#F8F8FF')
plt.gcf().set_facecolor('#F8F8FF')

plt.tight_layout()
plt.show()
```
The violin plot above showcases the cherry blossom flowering date distribution by century (9th to 21st), measured in day-of-year (DOY). The dashed pink line represents the average flowering day of the year (~105 days). It was quite consistent until the 20th century (1900s), where we can see a large drop; this trend has also been carried into the 21st century, where it even dropped lower.

MODERN DATA

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

@@

# Filter out NaN values
filtered_modern = modern.dropna(subset=['flower_doy'])

# Create scatter plot with pink color
plt.scatter(filtered_modern['year'], filtered_modern['flower_doy'], alpha=0.7, color='#FF9ECD', 
            edgecolor='#E75480', s=50, label='Flowering Day')

# Add moving average line with light blue color
window_size = 5  # 5-year moving average
filtered_modern['moving_avg_doy'] = filtered_modern['flower_doy'].rolling(window=window_size).mean()
plt.plot(filtered_modern['year'], filtered_modern['moving_avg_doy'], color='#4BA3C3', linewidth=2.5, 
         label=f'{window_size}-Year Moving Avg')

# Customize the plot with pink/blue theme
plt.title('Cherry Blossom Flowering Day Throughout Modern Times', fontsize=16, color='#E75480')
plt.xlabel('Year', fontsize=12, color='#4BA3C3')
plt.ylabel('Day of Year (DOY)', fontsize=12, color='#4BA3C3')
plt.grid(True, alpha=0.3, color='#E0E0E0')
plt.xticks(color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.legend(framealpha=0.7)

# Create a secondary y-axis with date labels
ax = plt.gca()
ax2 = ax.twinx()

# Get the current y-limits from the primary axis
y_min, y_max = ax.get_ylim()

# Create properly aligned DOY-to-date mapping
# In a non-leap year:
# DOY 90 = March 31
# DOY 91 = April 1
# DOY 121 = May 1

doy_values = np.arange(int(y_min), int(y_max) + 1, 5)  # Create ticks every 5 days
date_labels = []

for doy in doy_values:
    if doy < 90:
        # DOYs below 90 (March 31)
        day_in_mar = doy - 59  # 59 = DOYs in Jan (31) + Feb (28)
        date_labels.append(f'Mar {day_in_mar}')
    elif doy < 120:
        # DOYs for April (90-119)
        day_in_apr = doy - 90 + 1
        date_labels.append(f'Apr {day_in_apr}')
    elif doy < 151:
        # DOYs for May (120-150)
        day_in_may = doy - 120 + 1
        date_labels.append(f'May {day_in_may}')
    else:
        # DOYs for June and beyond
        day_in_jun = doy - 151 + 1
        date_labels.append(f'Jun {day_in_jun}')

# Set the properly aligned ticks and labels
ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3', fontsize=14)
ax2.tick_params(axis='y', colors='#4F86C6')

plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, axis='y')

# Add pink/blue background touches
plt.gca().set_facecolor('#F8F8FF')  # Light background
plt.gcf().set_facecolor('#F8F8FF')

plt.tight_layout()
plt.show()
```
The scatterplot above presents the annual full flowering dates of cherry blossoms in Japan from 1950 to 2020, measured in day-of-year (DOY), where higher values represent later blooming. Each pink dot marks the full flowering day for a given year, while the solid blue line shows a 5-year moving average, smoothing short-term fluctuations to reveal broader temporal trends. Over the 70-year span, there is a clear downward trend, indicating progressively earlier flowering.

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

@@

# Add moving average line in light blue
modern['moving_avg'] = modern['avg_temp_march_c'].rolling(window=5).mean()  # Using 5-year window for modern data
plt.plot(modern['year'], modern['moving_avg'], color='#4BA3C3', linewidth=2.5, label='5-Year Moving Avg')

@@

# Add pink/blue background touches
plt.gca().set_facecolor('#F8F8FF')  # Light background
plt.gcf().set_facecolor('#F8F8FF')

plt.tight_layout()
plt.show()
```
This scatterplot shows the annual spring temperature in Japan from 1950 to 2020, measured in degrees Celsius (°C) relative to a baseline average. Each pink dot represents the average March temperature for a given year, and the solid blue line indicates a 5-year moving average to highlight underlying climate trends. The overall trajectory shows a clear increase in average March temperatures over the 70-year period, particularly after 1990, reflecting the impact of modern global climate change. This warming trend corresponds closely with the earlier flowering of cherry blossoms during the same time frame shown in the earlier figure. Together, both graphs reinforce the connection between rising spring temperatures and timing shifts in sakura blooming.

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

# histogram of flowering days
plt.figure(figsize=(12, 7))
plt.hist(modern['flower_doy'], bins=30, color='#FF9ECD', edgecolor='#E75480', alpha=0.7)
plt.title('Distribution of Flowering Days (Modern)', fontsize=18, color='#E75480')
plt.xlabel('Day of Year (DOY)', fontsize=14, color='#4BA3C3')
plt.ylabel('Frequency', fontsize=14, color='#4BA3C3')
plt.xticks(color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, color='#E0E0E0')
plt.gca().set_facecolor('#F8F8FF')  # Light background
plt.gcf().set_facecolor('#F8F8FF')
plt.tight_layout()
plt.show()
```
This histogram highlights the frequency distribution of sakura full flowering dates in modern Japan, measured in day-of-year (DOY), for the period between 1950 and 2020. Each bar represents how often flowering occurred within a specific DOY range. The distribution is roughly bell-shaped and unimodal, with the majority of flowering dates clustered between DOY 83 and DOY 95 (approximately March 24 to April 4), suggesting a central tendency in late March.

```{python}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: WRITE SOMETHING
#| fig-align: center

>>>>

Before the advent of modern technology, people understood themselves, their environments, and their cultures through the cycle of life and death. The Japanese tradition of *hanami* (loosely translated as 'flower-viewing') is an important cultural tradition that honors this perspective, being associated with the flowering birth of cherry blossom trees in the late spring. In the city of Kyoto in particular, local newspapers have recorded the dates of tree flowering since the late 19th century, but researchers like @Aono2010 have been able to corroborate flowering dates as far back as the year 812 CE, through the examination of historical records such as diaries, poetry and other written documents describing *hanami* parties throughout the ages. However, rising global temperatures have created anomalous climate conditions in recent years.  This highlights a need to understand the historical and modern links between temperature and cherry blossom bloom timing. Without it, cherished traditions like *hanami* may struggle to adapt to climate shifts, risking their disappearance as unique customs.

Work such as @Allen2014 analyzes cherry blossom blooming data over a relatively brief 29-year timespan. However, this narrow temporal window represents only a fraction of the documented flowering history, limiting their model's ability to capture the relationship between blossoms and the significant climate shifts of the past 150 years - shifts that are clearly visible in @Aono2010's more comprehensive dataset.

We also seek to answer the question of whether warmer temperatures lead to earlier blooming times. To achieve this, we instead analyse and model the nearly 1200-year record compiled by @Aono2010 in tandem with modern data compiled by the Japanese Meteorological Agency (JMA), collected and compiled together by @Cookson. @sec-lit-review does a broad review of existing work surrounding cherry blossoms, their significance, and efforts to understand their behaviour related to climate. @sec-data covers the data sources, methodology, and conducts an exploratory analysis of the dataset. @sec-model proposes a model of cherry blossom flowering day as a function of mean March temperature in Kyoto and lists its results. @sec-discussion critically examines the data and results, comes to several conclusions about the dataset, and proposes new areas of exploration. This analysis highlights the  consequential effects for wildlife that function around the timing of blooming *sakura*, the need for flexibility in planning *hanami* celebrations going forward, as well as the need for wider ranging data to inform more precise predictions of blooming days.

@@

```{python}
#| label: tbl-historical
#| tbl-cap: "Historical Cherry Blossom Blooming"

# Rename columns for better readability
historic_bloom_renamed = historic_bloom.rename(columns={
    'year': 'Year',
    'flower_date': 'Bloom Date',
    'flower_doy': 'Day of Year',
    'avg_temp_march_c': 'March Temp (°C)'
})

# Display the table without .style.set_caption()
display(historic_bloom_renamed.head())
```

@@

```{python}
#| label: tbl-modern
#| tbl-cap: "Modern Cherry Blossom Blooming"

@@

In our data exploration, the `matplotlib` [@matplotlib] and `seaborn` [@seaborn] packages were used to visualize and explore the data. The historical dataset was visualized and explored first.

```{python}
#| label: fig-hist-temp-yr-scatter
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# Scatterplot of historic temp vs years 
plt.figure(figsize=(10, 6))

# Scatter plot with pink points
plt.scatter(historic['year'], historic['avg_temp_march_c'], alpha=0.7, color='#FF9ECD', edgecolor='#E75480', s=50)

# Add moving average line
moving_avg = historic['avg_temp_march_c'].rolling(window=10).mean()
plt.plot(historic['year'], moving_avg, color='#5F3E2A', linewidth=2.5, label='10-Year Moving Avg')

# Add reference line for mean temperature
mean_temp = historic['avg_temp_march_c'].mean()
plt.axhline(y=mean_temp, color='skyblue', linestyle='--', label=f'Historical Mean ({mean_temp:.1f}°C)')

# Add labels and customize
plt.title('Historic Temperature vs Year', fontsize=16)
plt.xlabel('Year', fontsize=12)
plt.ylabel('Temperature (°C)', fontsize=12)
plt.grid(True, alpha=0.3)
plt.legend()

plt.tight_layout()
```

@fig-hist-temp-yr-scatter displays historical annual temperature records (in °C) from the year 800 CE to 2000 CE. Each pink dot represents the recorded temperature for a given year, while the solid teal line represents a 10-year moving average, smoothing out year-to-year fluctuations to highlight longer-term trends. The dashed horizontal line represents the historical mean temperature of 6.2°C. Notably, the data show considerable fluctuations over the centuries, with a relatively stable pattern through much of the pre-industrial era. However, a sharp and continuous rise in average temperature is observed beginning in the late 19th to early 20th century, suggesting significant warming recently. This trend aligns with broader patterns of climate change and may have implications for phenological events such as the timing of sakura (cherry blossom) flowering.

```{python}
#| label: fig-hist-bloomday-yr
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# Create simplified flowering day history plot
plt.figure(figsize=(10, 6))

@@

# Create scatter plot 
plt.scatter(filtered_historic['year'], filtered_historic['flower_doy'], 
            alpha=0.7, color='#FF9ECD', edgecolor='#E75480', s=40)

# Add moving average line
moving_avg = filtered_historic['flower_doy'].rolling(window=10).mean()
plt.plot(filtered_historic['year'], moving_avg, 
         color='#6E4B34', linewidth=2, label='10-Year Moving Avg')

# Add reference line for historical mean
mean_doy = filtered_historic['flower_doy'].mean()
plt.axhline(y=mean_doy, color='#87CEFA', linestyle='--', 
            label=f'Historical Mean ({mean_doy:.1f}, April 13-14)')

# Set up secondary y-axis for date labels
ax = plt.gca()
ax2 = ax.twinx()
y_min, y_max = ax.get_ylim()
y_min, y_max = int(y_min), int(y_max)

# Create date labels
doy_values = np.arange(y_min - (y_min % 10), y_max + 10, 10)  # Ticks every 10 days
date_labels = []
months = {3: 'Mar', 4: 'Apr', 5: 'May', 6: 'Jun'}

for doy in doy_values:
    # Convert DOY to month/day
    if doy <= 90:
        date_labels.append(f'{months[3]} {doy-59}')
    elif doy <= 120:
        date_labels.append(f'{months[4]} {doy-90+1}')
    elif doy <= 151:
        date_labels.append(f'{months[5]} {doy-120+1}')
    else:
        date_labels.append(f'{months[6]} {doy-151+1}')

# Apply secondary axis settings
ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3')

# Customize appearance
plt.title('Cherry Blossom Flowering Day Throughout History', fontsize=14)
plt.xlabel('Year')
plt.ylabel('Day of Year (DOY)')
plt.grid(True, alpha=0.3)
plt.legend()
plt.tight_layout()
```

@fig-hist-bloomday-yr showcases the annual full flowering dates of sakura from 800 to 2000 CE, plotted as day-of-year (DOY) values, where higher values indicate later blooming in the calendar year. Each pink dot corresponds to an individual year's observed full bloom date, and the solid brown line shows a 10-year moving average; the dashed horizontal line indicates the historical mean, which is 104.6. The data suggest that while flowering dates have fluctuated across centuries, a notable trend toward earlier blooming began in the late 19th and 20th centuries. This shift coincides with the period of modern climate warming seen in the corresponding temperature graph, suggesting that rising spring temperatures are advancing sakura bloom time.

```{python}
#| label: fig-hist-violin-plot  
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# Violin Plot 
plt.figure(figsize=(14, 8))

# Create century labels based on year
if 'century_label' not in historic.columns:
    century_num = (historic['year'] // 100 + 1).astype(int)
    historic['century_label'] = century_num.apply(lambda x: f"{x}{'th' if x not in [1, 2, 3, 21] else ['st', 'nd', 'rd', 'st'][x % 10 - 1 if x % 100 != 11 else 0]}")

# Define centuries in chronological order
century_order = [f"{i}{'th' if i not in [21] else 'st'}" for i in range(9, 22)]

# Create a color gradient from light blue to pink
colors = ['#ADD8E6', '#FFC0CB']  # Light blue to Light pink
cm = LinearSegmentedColormap.from_list('blue_to_pink', colors, N=len(century_order))
custom_palette = {century: cm(i/float(len(century_order)-1)) for i, century in enumerate(century_order)}

# Create violin plot
sns.violinplot(x='century_label', y='flower_doy', data=historic, 
               palette=custom_palette, inner='quartile', order=century_order)

# Add data points
sns.stripplot(x='century_label', y='flower_doy', data=historic, 
              color='#FF9ECD', alpha=0.3, jitter=True, size=4, order=century_order)

# Add mean reference line
mean_doy = historic['flower_doy'].mean()
plt.axhline(y=mean_doy, color='#E75480', linestyle='--', 
            label=f'Historical Mean ({mean_doy:.1f}, April 14)')

# Set titles and labels
plt.title('Cherry Blossom Flowering Day Distribution by Century', fontsize=18, color='#E75480')
plt.xlabel('Century', fontsize=14, color='#4BA3C3')
plt.ylabel('Day of Year (DOY)', fontsize=14, color='#4BA3C3')
plt.xticks(rotation=45, color='#4F86C6')

# Add secondary y-axis with calendar dates
ax = plt.gca()
ax2 = ax.twinx()
y_min, y_max = ax.get_ylim()
doy_values = range(int(y_min), int(y_max) + 1, 10)  # Ticks every 10 days

# Create date labels
date_labels = []
for doy in doy_values:
    if doy < 90:
        date_labels.append(f'Mar {doy - 59}')
    elif doy < 120:
        date_labels.append(f'Apr {doy - 90 + 1}')
    elif doy < 151:
        date_labels.append(f'May {doy - 120 + 1}')
    else:
        date_labels.append(f'Jun {doy - 151 + 1}')

ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3', fontsize=14)

plt.legend()
plt.grid(True, alpha=0.3, axis='y')
plt.tight_layout()
```

@fig-hist-violin-plot showcases the cherry blossom flowering date distribution by century (9th to 21st), measured in day-of-year (DOY). The dashed pink line represents the average flowering day of the year (~105 days). It was quite consistent until the 20th century (1900s), where we can see a large drop; this trend has also been carried into the 21st century, where it dropped further.

A common thread emerges across all of the plots - warmer temperatures as time goes on has led to cherry blossoms blooming earlier and earlier every year.Furthermorethe significant shifting blooming days relative to temperature point to this historical data being ineffective in understanding the present relationship between temperature and blooming day. Since blooming days and temperatures have shifted so much in the last 150 years, examining more modern data is more effective in uncovering current trends. Given this, another more current dataset from @Cookson was examined next.

```{python}
#| label: fig-modern-year-temp 
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

@@

# Create scatter plot
plt.scatter(modern['year'], modern['flower_doy'], 
            alpha=0.7, color='#FF9ECD', edgecolor='#E75480', s=50)

# Add moving average line
moving_avg = modern['flower_doy'].rolling(window=5).mean()
plt.plot(modern['year'], moving_avg, color='#6E4B34', linewidth=2.5, 
         label='5-Year Moving Avg')

# Customize plot appearance
plt.title('Cherry Blossom Flowering Day Throughout Modern Times', fontsize=16, color='#E75480')
plt.xlabel('Year', fontsize=12, color='#4BA3C3')
plt.ylabel('Day of Year (DOY)', fontsize=12, color='#4BA3C3')
plt.grid(True, alpha=0.3)

# Create secondary y-axis with date labels
ax = plt.gca()
ax2 = ax.twinx()
y_min, y_max = ax.get_ylim()
doy_values = np.arange(int(y_min), int(y_max) + 1, 5)
date_labels = []

for doy in doy_values:
    if doy < 90:
        date_labels.append(f'Mar {doy-59}')
    elif doy < 120:
        date_labels.append(f'Apr {doy-90+1}')
    elif doy < 151:
        date_labels.append(f'May {doy-120+1}')
    else:
        date_labels.append(f'Jun {doy-151+1}')

ax2.set_ylim(y_min, y_max)
ax2.set_yticks(doy_values)
ax2.set_yticklabels(date_labels)
ax2.set_ylabel('Approximate Date', color='#4BA3C3')

plt.legend()
plt.tight_layout()
```

@fig-modern-year-temp above presents the annual full flowering dates of cherry blossoms in Japan from 1950 to 2020, measured in day-of-year (DOY), where higher values represent later blooming. Each pink dot marks the full flowering day for a given year, while the solid blue line shows a 5-year moving average, smoothing short-term fluctuations to reveal broader temporal trends. Over the 70-year span, there is a clear downward trend, indicating progressively earlier flowering.

```{python}
#| label: fig-modern-bloomday-yr
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

@@

# Add moving average line in light blue
modern['moving_avg'] = modern['avg_temp_march_c'].rolling(window=5).mean()  # Using 5-year window for modern data
plt.plot(modern['year'], modern['moving_avg'], color='#6E4B34', linewidth=2.5, label='5-Year Moving Avg')

@@

plt.tight_layout()
```

@fig-modern-bloomday-yr shows the annual spring temperature in Japan from 1950 to 2020, measured in degrees Celsius (°C) relative to a baseline average. Each pink dot represents the average March temperature for a given year, and the solid blue line indicates a 5-year moving average to highlight underlying climate trends. The overall trajectory shows a clear increase in average March temperatures over the 70-year period, particularly after 1990, reflecting the impact of modern global climate change. This warming trend corresponds closely with the earlier flowering of cherry blossoms during the same time frame shown in the earlier figure. Together, both graphs reinforce the connection between rising spring temperatures and timing shifts in sakura blooming.

```{python}
#| label: fig-modern-temps-hist
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# histogram of temperatures
plt.figure(figsize=(12, 7))
plt.hist(modern['avg_temp_march_c'], bins=30, color='#FF9ECD', edgecolor='#E75480', alpha=0.7)
plt.title('Distribution of Average March Temperatures (Modern)', fontsize=18, color='#E75480')
plt.xlabel('Temperature (°C)', fontsize=14, color='#4BA3C3')
plt.ylabel('Frequency', fontsize=14, color='#4BA3C3')
plt.xticks(color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, color='#E0E0E0')
plt.tight_layout()
```

@fig-modern-temps-hist above presents the frequency distribution of average March temperatures in Japan from 1950 to 2020. The temperature values are plotted on the x-axis in degrees Celsius (°C). The distribution is centered around -1.5°C, with a wider spread toward colder extremes and a thinner tail extending into positive temperatures. While most March temperatures during this period fall below freezing, the presence of warmer years, particularly in the latter decades, indicates a shift in climate norms. These warming trends align with observed shifts in cherry blossom phenology, where earlier flowering is increasingly correlated with milder spring conditions.

```{python}
#| label: fig-modern-flowering-days-hist
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# histogram of flowering days
plt.figure(figsize=(12, 7))
plt.hist(modern['flower_doy'], bins=30, color='#FF9ECD', edgecolor='#E75480', alpha=0.7)
plt.title('Distribution of Flowering Days (Modern)', fontsize=18, color='#E75480')
plt.xlabel('Day of Year (DOY)', fontsize=14, color='#4BA3C3')
plt.ylabel('Frequency', fontsize=14, color='#4BA3C3')
plt.xticks(color='#4F86C6')
plt.yticks(color='#4F86C6')
plt.legend(framealpha=0.7)
plt.grid(True, alpha=0.3, color='#E0E0E0')
plt.tight_layout()
plt.show()
```

@fig-modern-flowering-days-hist highlights the frequency distribution of sakura full flowering dates in modern Japan, measured in day-of-year (DOY), for the period between 1950 and 2020. Each bar represents how often flowering occurred within a specific DOY range. The distribution is roughly normally distributed, with the majority of flowering dates clustered between DOY 83 and DOY 95 (approximately March 24 to April 4), suggesting a central tendency in late March.

Exploring the data collected in the last 60 years has highlighted it's similarity to the  historical dataset. As temperatures have gotten warmer, cherry blossoms in Kyoto have tended to bloom earlier and earlier as the years have progressed. Although the temperatures in the modern dataset tend to be a bit cooler on average than the ones noted in the historical dataset, the clear negative relationship between temperature and blooming day endures. To be able to more exactly quantify this relationship, however, we construct a model.
